---
title: "Assignment: Course Project 1"
output:
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

### Downloading the dataset

```{r echo=TRUE}
compressedDatasetFileName <- 'dataset.zip'
if (!file.exists(compressedDatasetFileName)) {
  url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
  download.file(url, method = 'curl', destfile = compressedDatasetFileName)
}
```

### Extract dataset from zip archive

```{r echo=TRUE}
extractedDatasetFileName <- 'activity.csv'
if (!file.exists(extractedDatasetFileName)) {
  unzip(extractedDatasetFileName)
}
```

### load dataset from file

```{r echo=TRUE}
dataset <- read.csv(extractedDatasetFileName)
```

## What is mean total number of steps taken per day?

### Calculate the total number of steps taken per day

```{r echo=TRUE}
library(dplyr)
stepsPerDay <- select(dataset, steps, date) %>% group_by(date) %>% summarise(steps = sum(steps))
head(stepsPerDay)
```

### Make a histogram of the total number of steps taken each day

```{r qplot}
library(ggplot2)
stepsPerDay <- select(dataset, steps, date) %>% group_by(date) %>% summarise(steps = sum(steps, na.rm=TRUE))
ggplot(stepsPerDay, aes(steps)) + geom_histogram()
```

### Calculate and report the mean and median of the total number of steps taken per day

```{r echo=TRUE}
meanStepsPerDay <- mean(stepsPerDay$steps, na.rm = TRUE)
paste('mean of the total number of steps taken per day is', meanStepsPerDay)
medianStepsPerDay <- median(stepsPerDay$steps, na.rm = TRUE)
paste('median of the total number of steps taken per day is', medianStepsPerDay)
```

## What is the average daily activity pattern?

### Make a time series plot

```{r echo=TRUE}
stepsPerInterval <- select(dataset, steps, interval) %>% group_by(interval) %>% summarise(steps = mean(steps, na.rm=TRUE))
head(stepsPerInterval)
plot(stepsPerInterval$interval, stepsPerInterval$steps, type='l', main = 'Average daily activity pattern', xlab = 'Interval', ylab = 'Average steps')
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE}
maxNumberOfSteps <- stepsPerInterval[stepsPerInterval$steps == max(stepsPerInterval$steps),]
paste('The interval that contains the maximum number of steps is', maxNumberOfSteps$interval, 'and contains', maxNumberOfSteps$steps, 'steps')
```

## Imputing missing values

### Calculate and report the total number of in the dataset
```{r echo=TRUE}
paste('there are missing values', nrow(filter(dataset, is.na(steps))))
```

### Devise a strategy for filling in all of the missing values in the dataset

I'm going to apply the mean of the day and replace the NA with it. If no value for the day and result is NaN the mean is replaced by 0.

### Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r echo=TRUE}
newDataset = dataset[,]
datesWithNa <- newDataset %>% filter(is.na(steps)) %>% select(date) %>% unique()

for (date in datesWithNa$date) {
  dateMean <- mean(newDataset[newDataset$date == date, "steps"], na.rm = TRUE)
  if (is.nan(dateMean)) {
    dateMean <- 0
  }
  newDataset[newDataset$date == date & is.na(newDataset$steps), "steps"] <- dateMean
}
```

### Make a histogram of the total number of steps taken each day

```{r }
stepsPerDay <- select(newDataset, steps, date) %>% group_by(date) %>% summarise(steps = sum(steps, na.rm=TRUE))
ggplot(stepsPerDay, aes(steps)) + geom_histogram()
```

```{r echo=TRUE}
meanStepsPerDay <- mean(stepsPerDay$steps, na.rm = TRUE)
paste('mean of the total number of steps taken per day is', meanStepsPerDay)
medianStepsPerDay <- median(stepsPerDay$steps, na.rm = TRUE)
paste('median of the total number of steps taken per day is', medianStepsPerDay)
```

## Are there differences in activity patterns between weekdays and weekends?

### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” 
```{r }
newDataset$date <- as.Date(newDataset$date)
weekend = c('Saturday', 'Sunday')
newDataset <- newDataset %>% mutate(weekday= ifelse(weekdays(date) %in% weekend, 'weekend', 'weekday'))
head(newDataset)
newDataset$weekday <- factor(newDataset$weekday)
class(newDataset$weekday)
stepsPerInterval <- select(newDataset, steps, weekday, interval) %>% group_by(weekday, interval ) %>% summarise(steps = mean(steps, na.rm=TRUE))
head(stepsPerInterval)
```

### Make a panel plot containing a time series plot of the 5-minute intervage and the average number of steps
```{r }
qplot(interval, steps, data = stepsPerInterval, facets = weekday ~ . , geom = 'line', xlab = 'Interval', ylab = 'Number of steps')
```
